---
title: "MSA Job Creation"
author: "John Mwangi"
date: "10/26/2021"
output: html_document
---

# Objective

Time series model for determining number of job opportunities in Mombasa County

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Packages

```{r message=FALSE, warning=FALSE}
rm(list = ls())

#load(file = "26_Oct.RData")

library(tidyverse)
library(readxl)
library(forecast)
library(writexl)

theme_set(theme_light())
```

# Data importation

## KE Business Registrations

```{r}
ke_business_reg <-
read_excel(path = "../Data requirements/KE_business_registrations.xlsx") %>% 
  janitor::clean_names() %>% 
  select(year, adjusted) %>% 
  filter(!is.na(adjusted)) %>% 
  mutate(adjusted = round(adjusted)) %>% 
  rename(business_reg = adjusted) %>% 
  mutate(year = as.numeric(year))

ke_business_reg
```

## KE Inflation Rates

```{r}
ke_inflation_rates <-
read_excel(path = "../Data requirements/KE_inflation_rates.xls",
           sheet = "2021") %>% 
  select(Year, Value) %>% 
  janitor::clean_names() %>% 
  rename(inflation_rate = value) %>% 
  mutate(year = as.numeric(year))

ke_inflation_rates
```

## KE Employment Ratios

```{r}
ke_employment_ratios <-
read_excel(path = "../Data requirements/KE_employment_ratios.xls", 
           skip = 3) %>% 
  janitor::clean_names() %>% 
  filter(country_name=="Kenya") %>% 
  select(country_name, starts_with("x")) %>% 
  pivot_longer(cols = starts_with("x"),
               names_to = "year",
               values_to = "employment_ratio") %>% 
  mutate(year = str_remove(string = year, pattern = "x"),
         year = as.numeric(year),
         employment_ratio = employment_ratio/100) %>% 
  select(-country_name)

ke_employment_ratios
```

## KE GDP Annual Growth

```{r}
ke_gdp <-
read_excel(path = "../Data requirements/KE_gdp.xls",
           sheet = "2021") %>% 
  janitor::clean_names() %>% 
  rename(gdp_growth = value) %>% 
  mutate(gdp_growth = gdp_growth/100) %>% 
  mutate(year = as.numeric(year))

ke_gdp
```

## Mombasa Population

Projections of Mombasa population until 2030. We'll use the UN population estimates as they are more recent but age group estimates from county sources. Assume age group sizes are changing at a constant rate.

Ref:
https://worldpopulationreview.com/world-cities/mombasa-population
https://www.mombasa.go.ke/wp-content/uploads/2018/11/MSA-FINAL-CIDP-2018-22.pdf (page 12)

```{r}
msa_population <-
read_csv(file = "../Data requirements/MSA_population.csv") %>% 
  janitor::clean_names() %>% 
  arrange(year) %>% 
  mutate(yrs_15_64 = parse_number(yrs_15_64),
         yrs_15_64 = yrs_15_64/100) %>% 
  mutate(working_pop = round(yrs_15_64*population)) %>% 
  select(-starts_with("x"))

msa_population
```

## KE Unemployment Rates

```{r}
ke_unemployment_rate <-
read_excel(path = "../Data requirements/KE_unemployment_rates.xls", 
           skip = 3) %>%
  janitor::clean_names() %>% 
  filter(country_name=="Kenya") %>% 
  select(country_name, starts_with("x")) %>% 
  pivot_longer(cols = starts_with("x"),
               names_to = "year",
               values_to = "unemployment_rate") %>% 
  mutate(year = str_remove(string = year, pattern = "x")) %>% 
  filter(!is.na(unemployment_rate)) %>% 
  select(-country_name) %>% 
  mutate(unemployment_rate = unemployment_rate/100) %>% 
  mutate(year = as.numeric(year))
ke_unemployment_rate
```

# Business Registrations Forecast

**Assumption:** Business registrations is a factor of National GDP.

National GDP was chosen since it is the only one that has forecasted values.

## ARIMAX model

```{r}
gdp_reg <-
ke_business_reg %>%
  full_join(ke_gdp, by = "year") %>%
  arrange(year) %>% 
  filter(!is.na(gdp_growth), !is.na(business_reg))

gdp_reg
```

```{r}
reg_model <- auto.arima(y = gdp_reg$business_reg, xreg = gdp_reg$gdp_growth)

reg_model
```

## Forecasted values

Forecast business registrations from 2022 to 2030. It is estimated that GDP will grow at 4.4% between 2020-2030.

```{r}
forecast(object = reg_model, 
         h = 9, 
         xreg = rep(0.044, times = 9)) %>% 
  plot()


reg_forecast <-
forecast(object = reg_model, 
         h = 9, 
         xreg = rep(0.044, times = 9)) %>% 
  as_tibble() %>% 
  janitor::clean_names() %>% 
  rename(business_reg = point_forecast) %>% 
  mutate(year = 2022:2030) %>% 
  bind_rows(ke_business_reg) %>% 
  arrange(year) %>% 
  relocate(year, .before = everything()) %>% 
  mutate(business_reg = round(business_reg))

reg_forecast
```

## CAGR

```{r}
CAGR_formula <- function(FV, PV, yrs) {
cagr <- ((FV/PV)^(1/yrs)-1)
return(cagr)
}

reg_cagr <- CAGR_formula(FV = 66120, PV = 46387, yrs = 2030-2022)
reg_cagr <- scales::percent(reg_cagr, accuracy = 0.01)
reg_cagr
```

## Plotted results

```{r}
reg_forecast %>% 
  ggplot(aes(x = year, y = business_reg)) +
  geom_line() +
  geom_ribbon(aes(ymin = lo_95, ymax = hi_95, alpha = 0.3), 
              fill = "grey70", 
              show.legend = FALSE) +
  scale_y_continuous(labels = scales::comma_format()) +
  scale_x_continuous(breaks = seq(2000, 2030, 5), minor_breaks = NULL) +
  expand_limits(y = 0, x = 2000) +
  labs(title = "Growth of Business Registrations in Kenya [2002-2030]",
       subtitle = glue::glue("Projected to grow at a CAGR of {reg_cagr} from 2022-2030"),
       x = NULL,
       y = "Number of Business Registrations")
```

# Unemployment Rate Forecast

## ARIMAX model

```{r}
reg_forecast
ke_gdp
ke_inflation_rates
ke_unemployment_rate
unemp_data <-
full_join(x = ke_gdp, y = reg_forecast, by = "year") %>% 
  full_join(ke_inflation_rates, by = "year") %>% 
  full_join(ke_unemployment_rate, by = "year") %>% 
  arrange(year) %>% 
  select(-starts_with(c("lo", "hi"))) %>% 
  relocate(unemployment_rate, .after = year) %>% 
  filter(complete.cases(.))
unemp_data
```


```{r}
unemp_data %>% 
  select(-year, -unemployment_rate) %>% 
  as.matrix()
```


```{r}
unemp_model <-
auto.arima(y = unemp_data$unemployment_rate, 
           xreg = unemp_data %>%
             select(-year, -unemployment_rate) %>%
             as.matrix())
unemp_model
```

## Forecasted values

Forecast unemployment rates from 2021-2030

```{r}
unemp_newdata <-
full_join(x = ke_gdp, y = reg_forecast, by = "year") %>% 
  full_join(ke_inflation_rates, by = "year") %>% 
  full_join(ke_unemployment_rate, by = "year") %>% 
  arrange(year) %>% 
  select(-starts_with(c("lo", "hi"))) %>% 
  relocate(unemployment_rate, .after = year) %>% 
  filter(year>=2021) %>% 
  select(-year, -unemployment_rate) %>% 
  as.matrix()
unemp_newdata
```

```{r}
forecast(object = unemp_model, h = 10, xreg = unemp_newdata) %>% plot()
unemp_forecast <-
forecast(object = unemp_model, h = 10, xreg = unemp_newdata) %>% 
  as_tibble() %>% 
  janitor::clean_names() %>% 
  mutate(year = 2021:2030) %>% 
  relocate(year, .before = everything()) %>% 
  rename(unemployment_rate = point_forecast)
unemp_forecast
```


```{r}
unemp_complete <-
full_join(x = ke_gdp, y = reg_forecast, by = "year") %>% 
  full_join(ke_inflation_rates, by = "year") %>% 
  full_join(ke_unemployment_rate, by = "year") %>% 
  arrange(year) %>% 
  select(-starts_with(c("lo", "hi"))) %>% 
  relocate(unemployment_rate, .after = year) %>% 
  left_join(unemp_forecast, by = "year") %>% 
  mutate(unemployment_rate = coalesce(unemployment_rate.x, unemployment_rate.y)) %>% 
  select(-unemployment_rate.x, -unemployment_rate.y) %>% 
  select(-contains("80")) %>% 
  arrange(year) %>% 
  filter(year>=2002)
unemp_complete
```

## Growth rate

```{r}
unemp_rates <-
unemp_complete %>% 
  filter(year>=2021) %>% 
  pull(unemployment_rate)
unemp_rates %>% 
  mean() %>% 
  scales::percent(accuracy = 0.01)
scales::percent(unemp_rates[length(unemp_rates)] - unemp_rates[1], accuracy = 0.01)
```

## Plotted results

```{r}
unemp_complete %>% 
  ggplot(aes(x = year, y = unemployment_rate)) +
  geom_line() +
  geom_ribbon(aes(ymin = lo_95, ymax = hi_95, alpha = 0.3), 
              fill = "grey70", 
              show.legend = FALSE) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_x_continuous(breaks = seq(2000, 2030, 5), minor_breaks = NULL) +
  geom_text(aes(label = scales::percent(unemployment_rate, accuracy = 0.01)), 
            check_overlap = TRUE) +
  expand_limits(x = 2000) +
  labs(title = "Unemployment Rates in Kenya [2002-2030]",
       subtitle = "Unemployment is expected to reduce 2.68% in 2030",
       x = NULL,
       y = "Unemployment Rate")
```

# Employment Ratio Forecast

## ARIMAX model

```{r}
reg_forecast
ke_gdp
ke_inflation_rates
ke_employment_ratios

emp_data <-
full_join(x = ke_gdp, y = reg_forecast, by = "year") %>% 
  full_join(ke_inflation_rates, by = "year") %>% 
  full_join(ke_employment_ratios, by = "year") %>% 
  arrange(year) %>% 
  select(-starts_with(c("lo", "hi"))) %>% 
  relocate(employment_ratio, .after = year) %>% 
  filter(complete.cases(.))

emp_data
```


```{r}
emp_data %>% 
  select(-year, employment_ratio) %>% 
  as.matrix()
```


```{r}
emp_model <-
auto.arima(y = emp_data$employment_ratio, 
           xreg = emp_data %>%
             select(-year, -employment_ratio) %>%
             as.matrix())

emp_model
```

## Forecasted values

Forecast unemployment rates from 2021-2030

```{r}
emp_newdata <-
full_join(x = ke_gdp, y = reg_forecast, by = "year") %>% 
  full_join(ke_inflation_rates, by = "year") %>% 
  full_join(ke_employment_ratios, by = "year") %>% 
  arrange(year) %>% 
  select(-starts_with(c("lo", "hi"))) %>% 
  relocate(employment_ratio, .after = year) %>% 
  filter(year>=2021) %>% 
  select(-year, -employment_ratio) %>% 
  as.matrix()

emp_newdata
```

```{r}
forecast(object = emp_model, h = 10, xreg = emp_newdata) %>% plot()

emp_forecast <-
forecast(object = emp_model, h = 10, xreg = emp_newdata) %>% 
  as_tibble() %>% 
  janitor::clean_names() %>% 
  mutate(year = 2021:2030) %>% 
  relocate(year, .before = everything()) %>% 
  rename(employment_ratio = point_forecast)

emp_forecast
```


```{r}
emp_complete <-
full_join(x = ke_gdp, y = reg_forecast, by = "year") %>% 
  full_join(ke_inflation_rates, by = "year") %>% 
  full_join(ke_employment_ratios, by = "year") %>% 
  arrange(year) %>% 
  select(-starts_with(c("lo", "hi"))) %>% 
  relocate(employment_ratio, .after = year) %>% 
  left_join(emp_forecast, by = "year") %>% 
  mutate(employment_ratio = coalesce(employment_ratio.x, employment_ratio.y)) %>% 
  select(-employment_ratio.x, -employment_ratio.y) %>% 
  select(-contains("80")) %>% 
  arrange(year) %>% 
  filter(year>=2002)

emp_complete
```

## Growth rate

```{r}
emp_ratios <-
emp_complete %>% 
  filter(year>=2020) %>% 
  pull(employment_ratio)

emp_ratios %>% 
  mean() %>% 
  scales::percent(accuracy = 0.01)

emp_2030 <- scales::percent(emp_ratios[length(emp_ratios)], accuracy = 0.01)
emp_2020 <- scales::percent(emp_ratios[1], accuracy = 0.01)
emp_2030
```

## Plotted results

```{r}
emp_complete %>% 
  ggplot(aes(x = year, y = employment_ratio)) +
  geom_line() +
  geom_ribbon(aes(ymin = lo_95, ymax = hi_95, alpha = 0.3), 
              fill = "grey70", 
              show.legend = FALSE) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_x_continuous(breaks = seq(2000, 2030, 5), minor_breaks = NULL) +
  geom_text(aes(label = scales::percent(employment_ratio, accuracy = 0.01)), 
            check_overlap = TRUE) +
  expand_limits(x = 2000) +
  labs(title = "Employment Ratios in Kenya [2002-2030]",
       subtitle = glue::glue("Employment is expected to reduce to {emp_2030} by 2030 from {emp_2020} in 2020"),
       x = NULL,
       y = "Employment Ratio")
```

# Job Opportunity Estimates

## Projections

According to KNBS, Employment Ratio is an estimate of the number of employed persons in a demographic. It is calculated as `Employed Population/Population (15-64)`. Estimates are total number of job opportunities (existing + new).

```{r}
job_opportunities <-
emp_complete %>% 
  left_join(msa_population, by = "year") %>% 
  mutate(jobs_base = working_pop*employment_ratio,
         jobs_best = working_pop*hi_95,
         jobs_worst = working_pop*lo_95) %>% 
  mutate(across(.cols = starts_with("jobs_"), .fns = round))

job_opportunities
```

## Plotted results

```{r}
job_ops <-
job_opportunities %>% 
  filter(year>=2021) %>% 
  pull(jobs_base)

jobs_created <- job_ops[length(job_ops)] - job_ops[1]
jobs_created <- scales::comma(jobs_created)
jobs_created
```


```{r}
jobs_plot <-
job_opportunities %>% 
  filter(year>=2010) %>% 
  ggplot(aes(x = year, y = jobs_base)) +
  geom_col(aes(x = year, y = working_pop), fill = "#53868B") +
  geom_line() +
  geom_ribbon(aes(ymin = jobs_worst, ymax = jobs_best, alpha = 0.3), 
              fill = "grey70", 
              show.legend = FALSE) +
  scale_y_continuous(breaks = seq(0, 2000000, 250000), 
                     minor_breaks = NULL, 
                     labels = scales::comma_format()) +
  scale_x_continuous(breaks = seq(2000, 2030, 5), minor_breaks = NULL) +
  geom_text(aes(label = round(jobs_base/1000)), 
            check_overlap = TRUE, 
            size = 4) +
  labs(title = "Job Opportunities in Mombasa County ('000)",
       subtitle = glue::glue("{jobs_created} new jobs are expected to be created by 2030."),
       x = NULL,
       y = "Job Opportunities")

#plotly::ggplotly(p = jobs_plot)
jobs_plot
```


# Save results

```{r}
write_xlsx(x = list(job_opportunities = job_opportunities,
                    employment_ratios = emp_complete,
                    unemployment_rates = unemp_complete), 
           path = "./outputs/job_opportunities.xlsx")
```


```{r}
save.image(file = "27_Oct.RData")
```






